---
title: "Workshop 8 KEY"
subtitle: "insert your section time here"
author: "your name here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)  

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r packages-and-data}
# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  # file path
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  # specifying which sheet you want to read in
  sheet = "First Harvest"
)
```

Storing some colors to use in visualizations, and setting a common `ggplot` theme for all plots:

```{r storing-colors-and-theme}
# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

```{r data-cleaning}
# creating a clean data frame
drought_exp_clean <- drought_exp |> # starting with the drought_exp object
  # cleaning column names
  clean_names() |> 
  # making a column of full species names
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # making a column of full drought treatment names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  )) |> 
  # making sure that species_name is a factor and ordering levels
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # making sure that water_treatment is a factor and ordering levels
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting columns
  select(species_name, water_treatment, sla, total_g)
```

Double checking that everything worked by checking the structure:

```{r checking-structure}
str(drought_exp_clean)
```

And displaying some rows from the data frame:

```{r showing-data-frame-rows}
slice_sample(
  drought_exp_clean, # data frame
  n = 10 # showing 10 rows
)
```


# 3. Visualizing data

## What is the relationship between SLA and total mass?  

```{r sla}
#| fig-width: 10

# base layer: ggplot
ggplot(data = drought_exp_clean,
       aes(x = sla,
           y = total_g)) +
  # scatter plot
  geom_point() +
  # relabeling axes and adding a title
  labs(x = expression(Specific~leaf~area~(cm^2/g)),
       y = "Total mass (g)",
       title = "As SLA increases, total mass increases")
```

## What are the differences in total mass between water treatments?

```{r water-treatment}
#| fig-width: 10

# base layer: ggplot
ggplot(data = drought_exp_clean,
       aes(x = water_treatment,
           y = total_g,
           color = water_treatment)) +
  # first layer: jitter
  geom_jitter(width = 0.1,
              height = 0,
              shape = 21,
              alpha = 0.6) +
  # adding a summary to calculate mean mass in each water treatment
  stat_summary(geom = "point",
               fun = mean,
               size = 4) +
  # manually setting colors
  scale_color_manual(values = c("Well watered" = ww_col,
                                "Drought stressed" = ds_col)) +
  # changing axis labels and title
  labs(x = "Water treatment",
       y = "Total mass (g)",
       title = "Average total mass is higher in well-watered treatment than drought stressed treatment") +
  # taking out legend
  theme(legend.position = "none")
```

## What are the differences in total mass between species?

```{r species}
#| fig-width: 10

# base layer: ggplot
ggplot(data = drought_exp_clean,
       aes(x = species_name,
           y = total_g,
           color = species_name)) +
  # first layer: jitter
  geom_jitter(width = 0.1,
              height = 0,
              shape = 21,
              alpha = 0.6) +
  # summary to calculate mean mass for each species
  stat_summary(geom = "point",
               fun = mean,
               size = 4) +
  # making sure that axis labels wrap nicely
  scale_x_discrete(labels = label_wrap(10)) +
  # setting colors manually
  scale_color_manual(values = c(
    "Lotus scoparius" = lotsco_col,
    "Penstemon centranthifolius" = pencen_col,
    "Salvia leucophylla" = salleu_col,
    "Encelia californica" = enccal_col,
    "Stipa pulchra" = stipul_col,
    "Eschscholzia californica" = esccal_col,
    "Grindelia camporum" = gricam_col
  )) +
  # labelling axes and title
  labs(x = "Species",
       y = "Total mass (g)",
       title = "Species differ in mean mass") +
  # getting rid of legend
  theme(legend.position = "none")
```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|:------------:|:---:|:---------------:|:-------:|-----------------------------|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r model-fitting}
# model 0: null model
model0 <- lm(
  total_g ~ 1, # formula
  data = drought_exp_clean # data frame
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula
  data = drought_exp_clean # data frame
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)

# model 5: SLA
model5 <- lm(total_g ~ sla,
             data = drought_exp_clean)

# model 6: water treatment
model6 <- lm(total_g ~ water_treatment,
             data = drought_exp_clean)

# model 7: species
model7 <- lm(total_g ~ species_name,
             data = drought_exp_clean)
```

## Model diagnostics

```{r model-diagnostics}
par(mfrow = c(2,2))
plot(model1)
plot(model2)
plot(model3)
plot(model4) # residuals look ok for best model
plot(model5)
plot(model6)
plot(model7)
```

## Model selection

```{r model-selection}
AICc(model1,
     model2,
     model3,
     model4, # best model: water and species
     model5,
     model6,
     model7,
     model0) |> 
  # arranging output in descending order of AIC
  arrange(AICc)
```

Look at the model summary:

```{r model-summary}
summary(model4)
# reference level for water_treatment: Drought stressed
# reference level for species: Lotus scoparius
```

## Stop and think: what does this model mean?  

**What is the best model?**

The best model that predicts total plant mass (g) includes water treatment and species (F(7, 62) = 27.08, p < 0.001, $\alpha$ = 0.05), but not SLA.  

**How much variation in the response (total mass, in grams) does this model explain?**

This model explained 73% (adjusted R^2^) of the variation in total mass.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

On average, plant masses in drought stressed conditions were 0.12 $\pm$ 0.04 g less than plant masses in well watered conditions across species.  

Across water treatments, all species have different (and higher) masses than _Lotus scoparius_ except _Penstemon centranthifolius_.

# 5. Model predictions

```{r model-predictions}
model4_predictions <- ggpredict(
  model4, # model object
  terms = c("species_name", "water_treatment") # predictors
) |> 
  # treating this like a regular data frame
  # renaming the columns
  rename(species_name = x,
         water_treatment = group)
```

Before plotting, LOOK AT THE PREDICTION DATA FRAME using `View(model4_predictions)` or by clicking on the object in the Environment.

# 6. Final figure

```{r final-figure}
#| fig-width: 12
#| fig-height: 4

# base layer: ggplot
# using the drought_exp_clean object
ggplot(data = drought_exp_clean,
       aes(x = water_treatment,
           y = total_g,
           color = water_treatment)) +
  # first layer: jitter showing individual observations
  geom_jitter(width = 0.1,
              height = 0,
              shape = 21,
              alpha = 0.6) +
  # second layer: point range showing model predictions
  # note that this is appropriate because there are only categorical predictors
  # in the final model
  geom_pointrange(data = model4_predictions,
                  aes(y = predicted,
                      ymin = conf.low,
                      ymax = conf.high)) +
  # manually setting colors
  scale_color_manual(values = c("Well watered" = ww_col,
                                "Drought stressed" = ds_col)) +
  # wrapping x-axis labels
  scale_x_discrete(labels = label_wrap(10)) +
  # creating different panels for each species
  facet_wrap(~species_name, nrow = 1) +
  # labelling x- and y-axis
  labs(x = "Water treatment",
       y = "Total mass (g)") +
  # taking out the legend
  theme(legend.position = "none")
```

